# Comparison Tables {#comparison-tables}

## Purpose

This appendix provides side-by-side comparisons of major theories, models, and frameworks discussed throughout the book. The tables are intended as quick reference tools. They do not replace the detailed discussions in the chapters, but they help clarify how different positions answer the central question: did life give rise to consciousness, did consciousness make life possible, or did life and consciousness co-emerge?

The comparisons are organized around consciousness theories, origin-of-life models, transition points, philosophical frameworks, artificial consciousness, ethical implications, methodological strengths and limitations, and overall theory strengths and weaknesses.

The tables in this appendix use a dual rendering strategy. In HTML, wide tables appear in scrollable boxes. In PDF, wide tables are placed in landscape orientation so that columns remain readable.

```{r comparison-table-helper, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

render_comparison_table <- function(data, caption, widths, font_size = 8, landscape_pdf = TRUE, html_height = "600px") {
  if (knitr::is_latex_output()) {
    if (landscape_pdf) {
      cat("\\begin{landscape}
")
    }

    tbl <- kable(
      data,
      format = "latex",
      booktabs = TRUE,
      longtable = FALSE,
      caption = caption
    ) |>
      kable_styling(
        latex_options = c("hold_position", "scale_down"),
        font_size = font_size,
        full_width = FALSE
      )

    for (i in seq_along(widths)) {
      tbl <- tbl |> column_spec(i, width = widths[i])
    }

    print(tbl)

    if (landscape_pdf) {
      cat("\\end{landscape}
")
    }
  } else {
    kable(
      data,
      format = "html",
      caption = caption,
      escape = FALSE
    ) |>
      kable_styling(
        full_width = FALSE,
        bootstrap_options = c("striped", "hover", "condensed")
      ) |>
      scroll_box(width = "100%", height = html_height)
  }
}
```


---

## Consciousness Theories Compared

```{r comparison-consciousness-theories, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

consciousness_theories <- data.frame(
  Theory = c(
    "Global Workspace Theory",
    "Global Neuronal Workspace",
    "Integrated Information Theory",
    "Free Energy Principle",
    "Predictive Processing",
    "Attention Schema Theory",
    "Recurrent Processing Theory",
    "Higher-Order Theories",
    "Biological Naturalism",
    "Neural Darwinism",
    "Orch-OR",
    "Panpsychism",
    "Cosmopsychism",
    "Idealism",
    "Taheri’s T-Consciousness Framework",
    "Russellian Monism",
    "Dual-Aspect Monism",
    "Process Philosophy",
    "Emergentism"
  ),
  Type = c(
    "Scientific / cognitive",
    "Neuroscientific",
    "Scientific / philosophical",
    "Scientific / theoretical",
    "Scientific / cognitive",
    "Scientific / functional",
    "Neuroscientific",
    "Philosophical / cognitive",
    "Philosophical / biological",
    "Neurobiological",
    "Speculative / scientific",
    "Philosophical",
    "Philosophical",
    "Philosophical",
    "Consciousness-first / speculative",
    "Philosophical",
    "Philosophical",
    "Philosophical",
    "Philosophical / scientific"
  ),
  `Consciousness is...` = c(
    "Global availability or broadcast of information",
    "Neural ignition and sustained large-scale broadcasting",
    "Integrated information or intrinsic causal power",
    "Possibly linked to predictive self-maintenance",
    "Perception and cognition as prediction-error minimization",
    "The brain’s model of its own attention",
    "Recurrent or feedback neural processing",
    "Awareness of a mental state by a higher-order state",
    "A biological phenomenon caused by brain processes",
    "Dynamic re-entrant neural selection and integration",
    "Quantum objective reduction in microtubules",
    "A fundamental or ubiquitous feature of matter",
    "The universe as the primary conscious subject",
    "The primary or only reality",
    "Fundamental non-material consciousness expressed through consciousness fields",
    "The intrinsic nature of physical structure",
    "One reality with mental and physical aspects",
    "Experience or becoming as fundamental to reality",
    "An emergent property of complex organization"
  ),
  `Primary scope` = c(
    "Brains with global workspace architecture",
    "Brains with prefrontal-parietal and sensory integration",
    "Any system with sufficient integrated causal structure",
    "Living systems and active agents",
    "Brains and embodied agents",
    "Systems with attention schemas",
    "Nervous systems with recurrent circuits",
    "Systems capable of self-monitoring",
    "Biological brains",
    "Nervous systems with re-entrant processing",
    "Microtubule-containing biological systems",
    "All matter or all physical entities",
    "Cosmos as a whole; individuals as derivatives",
    "Everything within consciousness",
    "Matter, life, and mind as manifestations within T-Consciousness",
    "Matter as externally physical and internally experiential or proto-experiential",
    "All reality, viewed under two aspects",
    "Events, processes, and organisms",
    "Complex biological systems, especially brains"
  ),
  `Testable?` = c(
    "Yes",
    "Yes",
    "Partially",
    "Partially",
    "Yes, indirectly",
    "Yes",
    "Yes",
    "Partially",
    "Indirectly",
    "Partially",
    "Partially",
    "Not currently",
    "Not currently",
    "Not currently",
    "Difficult",
    "Difficult",
    "Difficult",
    "Not directly",
    "Indirectly"
  ),
  `Implication` = c(
    "Life first; requires specific neural architecture",
    "Life first; depends on advanced neural systems",
    "Co-emergence or panpsychist implications",
    "Co-emergence; roots of mind may lie in living regulation",
    "Life first or co-emergence",
    "Potentially substrate-independent",
    "Life first; requires feedback architecture",
    "Life first; requires self-representation",
    "Life first; requires biology",
    "Life first; emerges from neural organization",
    "May involve fundamental physics and life",
    "Consciousness first or co-emergence",
    "Consciousness first",
    "Consciousness first",
    "Consciousness first",
    "Consciousness first or co-emergence",
    "Co-emergence",
    "Co-emergence or consciousness first",
    "Life first"
  ),
  check.names = FALSE
)

render_comparison_table(
  consciousness_theories,
  caption = "Major consciousness theories compared.",
  widths = c("4.0cm", "3.5cm", "6.0cm", "6.0cm", "2.5cm", "5.0cm"),
  font_size = 7,
  landscape_pdf = TRUE,
  html_height = "650px"
)
```

This table shows that scientific theories tend to focus on mechanisms of access, integration, prediction, attention, recurrence, or biological organization. Philosophical and consciousness-first theories focus more directly on the ontological status of consciousness and the hard problem.

---

## Origin-of-Life Models and Consciousness Implications

```{r comparison-origin-life-models, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

origin_life_models <- data.frame(
  Model = c(
    "RNA World",
    "Metabolism-First",
    "Hydrothermal Vent Models",
    "Protocell Models",
    "Autocatalytic Sets",
    "Hypercycle Models",
    "Dissipative Structures",
    "Assembly Theory",
    "Autopoiesis",
    "Biosemiotic Models",
    "Consciousness-Enabled Life",
    "Taheri’s T-Consciousness View",
    "Artificial Life Models"
  ),
  `Key mechanism` = c(
    "Self-replicating RNA molecules capable of heredity and catalysis",
    "Self-sustaining chemical cycles precede genetic replication",
    "Energy gradients and mineral surfaces drive prebiotic chemistry",
    "Lipid compartments create boundaries and internal environments",
    "Molecular networks collectively catalyze their own formation",
    "Linked replicators support cooperative molecular evolution",
    "Energy flow maintains order far from equilibrium",
    "Molecular complexity measured by assembly steps",
    "Systems continuously produce and maintain themselves",
    "Life begins with signs, codes, and meaning-making",
    "Consciousness acts as organizing principle or condition",
    "Consciousness fields organize or manifest matter into living form",
    "Life-like systems created in digital, robotic, or synthetic media"
  ),
  `Implies agency?` = c(
    "Minimal",
    "Moderate",
    "Low to moderate",
    "Moderate",
    "Moderate",
    "Moderate",
    "Moderate",
    "Low",
    "High",
    "High",
    "High, but speculative",
    "High, but non-material and speculative",
    "Variable"
  ),
  `Consciousness relevance` = c(
    "Low; mainly explains replication and heredity",
    "Moderate; emphasizes self-maintaining organization",
    "Moderate; energy flow and structure may support self-organization",
    "Moderate to high; introduces self/non-self boundary",
    "Moderate; systemic self-maintenance resembles minimal agency",
    "Moderate; early coordination and information cycling",
    "Moderate; connects life to thermodynamic self-organization",
    "Low to moderate; useful for detecting life-like complexity",
    "High; directly connects life, boundary, agency, and meaning",
    "High; suggests life requires interpretation or significance",
    "High, but difficult to test",
    "High; treats life as an expression of prior consciousness",
    "Depends on embodiment, self-maintenance, and integration"
  ),
  `Implication for central question` = c(
    "Life first",
    "Life first or co-emergence",
    "Life first",
    "Life first or co-emergence",
    "Co-emergence-friendly",
    "Life first",
    "Co-emergence-friendly",
    "Life first",
    "Co-emergence",
    "Co-emergence or consciousness first",
    "Consciousness first",
    "Consciousness first",
    "Tests life-consciousness relationship"
  ),
  check.names = FALSE
)

render_comparison_table(
  origin_life_models,
  caption = "Origin-of-life models and consciousness implications.",
  widths = c("4.0cm", "7.0cm", "3.5cm", "6.5cm", "4.0cm"),
  font_size = 7,
  landscape_pdf = TRUE,
  html_height = "620px"
)
```

Origin-of-life models usually explain biological organization without directly explaining subjective experience. Models that emphasize autopoiesis, biosemiotics, and consciousness-enabled life are more relevant to the central question because they connect life with agency, meaning, or consciousness.

---

## Where Each Theory Places the Transition

```{r comparison-transition-points, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

transition_points <- data.frame(
  Theory = c(
    "Global Workspace Theory",
    "Global Neuronal Workspace",
    "Integrated Information Theory",
    "Free Energy Principle",
    "Predictive Processing",
    "Recurrent Processing Theory",
    "Attention Schema Theory",
    "Higher-Order Theory",
    "Biological Naturalism",
    "Orch-OR",
    "Panpsychism",
    "Cosmopsychism",
    "Idealism",
    "Taheri’s T-Consciousness Framework",
    "Emergentism",
    "Co-emergence"
  ),
  `Transition point` = c(
    "Emergence of global broadcast architecture",
    "Neural ignition across large-scale networks",
    "Presence or sufficient degree of integrated information",
    "Self-maintaining predictive regulation",
    "Deep embodied prediction and model integration",
    "Recurrent neural feedback",
    "Construction of a model of attention",
    "Mental states represented by higher-order states",
    "Biological brain processes with the right causal powers",
    "Quantum coherence and objective reduction in microtubules",
    "Consciousness or proto-consciousness always present",
    "Cosmic consciousness precedes individuals",
    "Consciousness is primary reality",
    "Transition is reframed as manifestation within T-Consciousness",
    "Sufficient biological or neural complexity",
    "Living self-organization becomes meaningful and integrated"
  ),
  `Gradual or sudden?` = c(
    "Threshold-like",
    "Threshold-like",
    "Gradual",
    "Gradual",
    "Gradual",
    "Threshold-like",
    "Threshold-like",
    "Threshold-like",
    "Threshold-like",
    "Event-based / threshold-like",
    "Not applicable",
    "Not applicable",
    "Not applicable",
    "Manifestation rather than emergence",
    "Threshold-like or gradual",
    "Gradual"
  ),
  `Earliest plausible conscious entity` = c(
    "Complex-brained animals",
    "Animals with advanced integrative brains",
    "Possibly any integrated system",
    "Potentially all living systems, depending on interpretation",
    "Organisms with complex predictive nervous systems",
    "Animals with recurrent sensory circuits",
    "Systems with attention self-models",
    "Animals capable of self-monitoring",
    "Biological animals with suitable brains",
    "Organisms with suitable microtubule organization",
    "All matter or all physical entities",
    "Universe as primary subject",
    "Universal consciousness",
    "Life as expression within consciousness fields",
    "Complex nervous systems",
    "Minimal living systems or early sentient organisms"
  ),
  `Main challenge` = c(
    "Explains access better than experience itself",
    "May exclude non-reportable or simpler consciousness",
    "Counterintuitive scope and measurement difficulties",
    "Free-energy minimization may not equal experience",
    "Prediction alone may not explain feeling",
    "Feedback may be necessary but not sufficient",
    "May explain reports of consciousness more than experience",
    "May over-intellectualize consciousness",
    "Does not specify exact mechanism",
    "Empirical support remains controversial",
    "Combination problem",
    "Decomposition problem",
    "Explaining physical regularity and individuation",
    "Mechanism and empirical testability remain unclear",
    "The hard transition remains unclear",
    "Needs sharper criteria"
  ),
  check.names = FALSE
)

render_comparison_table(
  transition_points,
  caption = "Where major theories place the transition to consciousness.",
  widths = c("4.2cm", "6.5cm", "3.8cm", "6.2cm", "5.5cm"),
  font_size = 7,
  landscape_pdf = TRUE,
  html_height = "650px"
)
```

This comparison shows why the “hard transition” remains difficult. Life-first theories must identify when biological or neural complexity becomes experience. Consciousness-first theories avoid this transition but inherit other problems, especially mechanism and testability.

---

## Philosophical Frameworks Compared

```{r comparison-philosophical-frameworks, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

philosophical_frameworks <- data.frame(
  Framework = c(
    "Substance Dualism",
    "Physicalism",
    "Reductive Materialism",
    "Non-Reductive Physicalism",
    "Functionalism",
    "Biological Naturalism",
    "Panpsychism",
    "Cosmopsychism",
    "Idealism",
    "Taheri’s T-Consciousness Framework",
    "Neutral Monism",
    "Dual-Aspect Monism",
    "Process Philosophy",
    "Enactivism",
    "Autopoietic Theory"
  ),
  `Mind-matter view` = c(
    "Mind and body are separate substances",
    "Mind is physical or fully dependent on physical processes",
    "Consciousness reduces to brain states",
    "Consciousness depends on but is not reducible to physical states",
    "Mind is functional organization",
    "Consciousness is biological and brain-based",
    "Mind-like properties exist in all matter",
    "Cosmic mind is fundamental",
    "Only mind or consciousness ultimately exists",
    "Consciousness is primary, non-material, and not produced by matter",
    "Mind and matter arise from a deeper neutral reality",
    "One reality has mental and physical aspects",
    "Reality is process, event, and becoming",
    "Mind arises through embodied action",
    "Life is self-producing organization"
  ),
  `Life-consciousness view` = c(
    "Separate but interacting questions",
    "Life first",
    "Life first",
    "Life first",
    "Substrate-independent",
    "Life first",
    "Consciousness first or co-emergence",
    "Consciousness first",
    "Consciousness first",
    "Consciousness first; life emerges through or within consciousness fields",
    "Co-emergence",
    "Co-emergence",
    "Co-emergence",
    "Co-emergence",
    "Co-emergence"
  ),
  Thinkers = c(
    "Descartes",
    "Smart, Armstrong, Dennett",
    "Identity theorists",
    "Various contemporary philosophers",
    "Putnam, Fodor",
    "Searle",
    "Strawson, Goff, Chalmers",
    "Shani, Goff, Nagasawa, Wager",
    "Berkeley, Kastrup",
    "Mohammad Ali Taheri",
    "Russell, James",
    "Spinoza, Pauli-Jung interpretations",
    "Whitehead",
    "Varela, Thompson, Di Paolo",
    "Maturana, Varela"
  ),
  Strength = c(
    "Preserves the reality of mind",
    "Fits scientific naturalism",
    "Parsimonious",
    "Allows higher-level properties",
    "Supports machine consciousness",
    "Takes biology and consciousness seriously",
    "Avoids emergence from non-conscious matter",
    "Avoids micro-combination problem",
    "Dissolves the hard problem",
    "Provides a direct consciousness-first account of life and mind",
    "Avoids strict dualism and reductionism",
    "Integrates mind and matter",
    "Integrates life, experience, and becoming",
    "Connects life, cognition, and meaning",
    "Strong account of living selfhood"
  ),
  Limitation = c(
    "Interaction problem",
    "Hard problem and explanatory gap",
    "May neglect subjective experience",
    "Risk of unclear causal status",
    "Chinese Room and simulation objections",
    "Mechanism remains unclear",
    "Combination problem",
    "Decomposition problem",
    "Explaining shared physical reality",
    "Mechanism and empirical testability remain unclear",
    "Nature of the neutral base unclear",
    "Difficult to test",
    "Metaphysically complex",
    "Consciousness boundary remains unclear",
    "Does not by itself prove consciousness"
  ),
  check.names = FALSE
)

render_comparison_table(
  philosophical_frameworks,
  caption = "Major philosophical frameworks compared.",
  widths = c("4.0cm", "6.0cm", "5.5cm", "4.5cm", "5.0cm", "5.0cm"),
  font_size = 6,
  landscape_pdf = TRUE,
  html_height = "650px"
)
```


Philosophical frameworks differ in what they treat as fundamental. Some begin with matter, some with consciousness, and others with a deeper neutral or process-based reality. These starting points strongly shape how each theory interprets life and consciousness.

---

## Life-First, Consciousness-First, and Co-Emergence Compared

```{r comparison-life-consciousness-positions, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

life_consciousness_positions <- data.frame(
  Position = c(
    "Life first",
    "Consciousness first",
    "Taheri’s T-Consciousness",
    "Co-emergence",
    "Strict physicalism",
    "Panpsychist continuum",
    "Biological naturalism",
    "Functionalism"
  ),
  `Basic claim` = c(
    "Life emerged first; consciousness evolved later",
    "Consciousness is fundamental; life emerges within it or from it",
    "Consciousness is a fundamental non-material reality; life manifests through consciousness fields",
    "Life and consciousness are deeply linked through self-organization and meaning",
    "Consciousness is fully physical and produced by matter",
    "Consciousness or proto-consciousness is widespread",
    "Consciousness requires biological brains",
    "Consciousness depends on functional organization"
  ),
  `Best explains` = c(
    "Neuroscience, evolution, animal consciousness, brain dependence",
    "The hard problem, contemplative traditions, idealist metaphysics",
    "Consciousness as organizing principle, non-material field ontology, life as expression",
    "Continuity between life, cognition, agency, and sentience",
    "Scientific parsimony and causal closure",
    "Avoids sudden emergence",
    "Brain dependence and living embodiment",
    "AI possibility and multiple realizability"
  ),
  `Struggles with` = c(
    "Hard problem, hard transition, subjective experience",
    "Mechanism, testability, combination or decomposition",
    "Scientific mechanism, testability, relation to physical chemistry",
    "Precision and falsifiability",
    "Phenomenal experience",
    "Over-attribution and combination problem",
    "AI and non-biological possibilities",
    "Simulation vs realization"
  ),
  `Ethical implication` = c(
    "Moral concern focused on sentient animals with suitable nervous systems",
    "Broader moral circle, possibly including nature or matter",
    "Broad moral and metaphysical implications; life is not merely material organization",
    "Layered moral concern across life, sentience, and ecosystems",
    "Narrower moral focus based on measurable sentience",
    "Very broad but graded moral concern",
    "Moral concern restricted to biological conscious beings",
    "Artificial systems may eventually matter morally"
  ),
  check.names = FALSE
)

render_comparison_table(
  life_consciousness_positions,
  caption = "Life-first, consciousness-first, and co-emergence positions compared.",
  widths = c("4.0cm", "6.0cm", "5.8cm", "5.5cm", "5.5cm"),
  font_size = 7,
  landscape_pdf = TRUE,
  html_height = "600px"
)
```

This table compares the three main families of answers developed throughout the book. Life-first theories are strongest scientifically, consciousness-first theories are strongest metaphysically, and co-emergence theories attempt to bridge life, cognition, and experience.

---

## Artificial Consciousness Models Compared

```{r comparison-artificial-consciousness, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

artificial_consciousness <- data.frame(
  View = c(
    "Computational functionalism",
    "Global Workspace view",
    "IIT-based view",
    "Attention Schema view",
    "Predictive processing view",
    "Biological naturalism",
    "Artificial life view",
    "Panpsychist view",
    "Idealist view"
  ),
  `Can AI be conscious?` = c(
    "Yes",
    "Possibly",
    "Possibly",
    "Possibly",
    "Possibly",
    "No, unless biological",
    "Possibly",
    "Possibly",
    "Possibly"
  ),
  `Required conditions` = c(
    "Correct computational organization",
    "Artificial global broadcast architecture",
    "High intrinsic integrated causal structure",
    "A model of attention and self-monitoring",
    "Embodied prediction, action, and self-modeling",
    "Biological causal powers of living brains",
    "Life-like self-maintenance, embodiment, adaptation",
    "Matter or systems with experiential aspects",
    "Artificial system as appearance within consciousness"
  ),
  `Main argument` = c(
    "Consciousness depends on function, not substrate",
    "Consciousness is global access",
    "Consciousness depends on integrated information",
    "Consciousness is an attention model",
    "Consciousness arises from active inference",
    "Simulation is not realization",
    "Consciousness may require life-like organization",
    "Consciousness is widespread",
    "Consciousness is fundamental"
  ),
  `Main objection` = c(
    "Symbol manipulation may not create understanding",
    "Access may not equal experience",
    "Digital systems may have low integrated causal structure",
    "May explain self-report, not feeling",
    "Prediction may not imply experience",
    "May be too restrictive",
    "Boundary between simulation and realization unclear",
    "Hard to test and specify",
    "Difficult to determine individual machine experience"
  ),
  check.names = FALSE
)

render_comparison_table(
  artificial_consciousness,
  caption = "Artificial consciousness models compared.",
  widths = c("5.0cm", "3.5cm", "6.0cm", "5.0cm", "5.5cm"),
  font_size = 8,
  landscape_pdf = TRUE,
  html_height = "550px"
)
```

Artificial consciousness tests whether biological life is necessary for experience. If consciousness depends only on functional organization, then AI could in principle be conscious. If consciousness depends on living embodiment, then artificial life may be more relevant than ordinary computation. If consciousness is fundamental, AI may raise a different question: how consciousness becomes individualized or expressed through artificial systems.

---

## Ethical Implications Compared

```{r comparison-ethical-implications, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

ethical_implications <- data.frame(
  `Theory / position` = c(
    "Human exceptionalism",
    "Mammal / bird sentience view",
    "Broad animal sentience view",
    "Basal cognition view",
    "Plant / fungal intelligence view",
    "Ecosystem ethics",
    "Panpsychism",
    "Biological naturalism",
    "Functionalism",
    "Taheri’s T-Consciousness view",
    "Precautionary approach",
    "Parsimony approach"
  ),
  `Likely moral circle` = c(
    "Humans only",
    "Mammals and birds, possibly some others",
    "Vertebrates plus many invertebrates",
    "Many living systems deserve respect",
    "Plants and fungi may have non-sentient value",
    "Ecosystems, species, rivers, habitats",
    "Very broad, possibly all matter",
    "Biological conscious organisms",
    "Biological and artificial conscious systems",
    "Broad metaphysical concern for life and consciousness",
    "Expands with uncertainty and evidence",
    "Narrower moral status"
  ),
  `Strongest ethical concern` = c(
    "Human rights and dignity",
    "Animal welfare",
    "Reducing suffering across animal life",
    "Respect for life and minimal agency",
    "Ecological humility and restraint",
    "Environmental protection and relational responsibility",
    "Avoiding dismissal of widespread interiority",
    "Animal and human sentience",
    "Possible AI welfare",
    "Life may have deeper consciousness-based significance",
    "Avoiding unrecognized suffering",
    "Avoiding unsupported attribution"
  ),
  Risk = c(
    "Excludes animals and non-verbal beings unfairly",
    "May exclude fish, insects, cephalopods, or crustaceans",
    "Difficult evidence boundaries",
    "May confuse cognition with sentience",
    "Anthropomorphism",
    "Collective moral status is difficult to define",
    "Moral inflation",
    "Excludes artificial consciousness",
    "Over-attribution to simulations",
    "Difficult to translate metaphysical concern into policy",
    "Practical limits and moral overload",
    "Under-recognition of sentience"
  ),
  check.names = FALSE
)

render_comparison_table(
  ethical_implications,
  caption = "Ethical implications of different consciousness positions.",
  widths = c("5.0cm", "6.0cm", "6.5cm", "5.5cm"),
  font_size = 8,
  landscape_pdf = TRUE,
  html_height = "600px"
)
```

The ethical implications of consciousness theories are not secondary. They shape how we treat animals, ecosystems, artificial systems, and life itself. The wider the theory places consciousness or sentience, the more expansive its moral implications become.

---

## Methodological Approaches Compared

```{r comparison-methodological-approaches, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

methodological_approaches <- data.frame(
  Method = c(
    "Behavioural observation",
    "Verbal report",
    "Neuroscience",
    "Brain imaging",
    "Anaesthesia studies",
    "Comparative cognition",
    "Basal cognition research",
    "Phenomenology",
    "Neurophenomenology",
    "Mathematical modelling",
    "Artificial intelligence experiments",
    "Synthetic biology / artificial life"
  ),
  `What it studies` = c(
    "Action, learning, response, adaptation",
    "Human subjective experience",
    "Brain activity and neural mechanisms",
    "Large-scale neural patterns",
    "Loss and recovery of consciousness",
    "Cross-species intelligence and behaviour",
    "Non-neural sensing and adaptation",
    "First-person structure of experience",
    "Links first-person and neural data",
    "Formal structure of theories",
    "Machine cognition and self-modeling",
    "Life-like organization"
  ),
  Strength = c(
    "Works across many organisms",
    "Direct access to reported experience",
    "Strong for humans and animals with brains",
    "Non-invasive evidence of conscious states",
    "Identifies necessary conditions",
    "Evolutionary perspective",
    "Expands cognition beyond brains",
    "Takes experience seriously",
    "Bridges subjective and objective methods",
    "Clarifies assumptions and predictions",
    "Tests substrate-independence",
    "Tests boundaries of life"
  ),
  Limitation = c(
    "Behaviour may be unconscious",
    "Limited to beings capable of report",
    "Correlation is not explanation",
    "Limited resolution and interpretation",
    "Does not explain experience itself",
    "Anthropomorphism risk",
    "Does not prove consciousness",
    "Subjective and difficult to verify",
    "Requires careful training and methods",
    "Model may not capture experience",
    "Intelligence may not equal sentience",
    "Consciousness remains difficult to infer"
  ),
  check.names = FALSE
)

render_comparison_table(
  methodological_approaches,
  caption = "Methodological approaches in consciousness research.",
  widths = c("5.0cm", "6.0cm", "5.5cm", "5.5cm"),
  font_size = 8,
  landscape_pdf = TRUE,
  html_height = "600px"
)
```

No single method is sufficient. Consciousness research requires behavioural, neural, phenomenological, comparative, formal, and ethical approaches. The central challenge is that consciousness is known directly from within but studied scientifically from without.

---

## Theory Strengths and Weaknesses at a Glance

```{r comparison-strengths-weaknesses, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

strengths_weaknesses <- data.frame(
  Theory = c(
    "GWT",
    "IIT",
    "FEP",
    "Predictive Processing",
    "AST",
    "RPT",
    "Biological Naturalism",
    "Orch-OR",
    "Panpsychism",
    "Cosmopsychism",
    "Idealism",
    "Russellian Monism",
    "Taheri’s T-Consciousness Framework",
    "Dual-Aspect Monism",
    "Process Philosophy",
    "Co-emergence"
  ),
  `Major strength` = c(
    "Explains conscious access and report",
    "Addresses intrinsic structure of experience",
    "Connects life, cognition, and regulation",
    "Explains perception as active inference",
    "Explains why systems model themselves as aware",
    "Offers a clear neural mechanism through feedback",
    "Respects biological basis of consciousness",
    "Connects consciousness to fundamental physics",
    "Avoids emergence from non-conscious matter",
    "Avoids micro-combination problem",
    "Makes consciousness primary and avoids the hard problem",
    "Bridges physics and consciousness through intrinsic nature",
    "Offers a clear consciousness-first model in which life and matter arise within a prior non-material consciousness order",
    "Avoids strict dualism and reductionism",
    "Integrates experience, life, and becoming",
    "Explains continuity between life and mind"
  ),
  `Major weakness` = c(
    "May not explain phenomenal experience",
    "Difficult to test and may imply too much consciousness",
    "Too broad if applied to all living systems",
    "Prediction alone may not produce experience",
    "May explain belief in consciousness rather than consciousness",
    "Feedback alone may not be sufficient",
    "Does not fully explain why biology feels",
    "Highly controversial and speculative",
    "Combination problem",
    "Decomposition problem",
    "Explaining physical regularity and shared reality",
    "Difficult to specify or test",
    "Requires clearer mechanisms, empirical criteria, and testable predictions",
    "Often too general",
    "Hard to formalize scientifically",
    "Needs sharper empirical criteria"
  ),
  check.names = FALSE
)

render_comparison_table(
  strengths_weaknesses,
  caption = "Theory strengths and weaknesses at a glance.",
  widths = c("5.0cm", "8.0cm", "8.0cm"),
  font_size = 9,
  landscape_pdf = TRUE,
  html_height = "600px"
)
```

---

## Summary

These comparison tables show that no single theory currently explains all dimensions of the life-consciousness relationship. Scientific theories often perform well on testability but struggle with subjective experience. Philosophical theories often address the hard problem more directly but struggle with mechanism and empirical evidence. Biological and enactive theories offer promising bridges, especially when they connect life, self-organization, meaning, and embodied agency.

The central pattern is clear:

* life-first theories are strongest on biology and neuroscience;
* consciousness-first theories are strongest on the hard problem and metaphysics;
* Taheri’s T-Consciousness framework provides a contemporary consciousness-first model centered on non-material consciousness fields;
* co-emergence theories are strongest on continuity between life, cognition, and mind;
* artificial consciousness tests whether life is necessary for experience;
* ethics forces theory into practical responsibility.

The unresolved question remains:

**Which framework can explain life, consciousness, experience, embodiment, meaning, and moral status without reducing one dimension to another?**
