# Consciousness Taxonomy {#consciousness-taxonomy}

## Purpose

This appendix provides a structured classification of the types, levels, dimensions, and theories of consciousness referenced throughout the book. The goal is not to impose a final definition of consciousness, but to give readers a map of the major ways consciousness can be classified.

Because the book moves across philosophy, biology, neuroscience, artificial intelligence, origin-of-life studies, and consciousness-first frameworks, the word “consciousness” is used in several related but distinct ways. This taxonomy helps clarify those uses.

This taxonomy is especially useful because different theories often use the same word in different ways. For example, consciousness may mean subjective experience, access to information, self-awareness, biological sentience, cosmic mind, or a fundamental non-material reality such as T-Consciousness. A clear taxonomy helps prevent these meanings from being collapsed into one another.

```{r taxonomy-table-helper, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

render_taxonomy_table <- function(data, caption, widths, font_size = 8, landscape_pdf = TRUE, html_height = "520px") {
  if (knitr::is_latex_output()) {
    if (landscape_pdf) {
      cat("\\begin{landscape}\n")
    }

    tbl <- kable(
      data,
      format = "latex",
      booktabs = TRUE,
      longtable = FALSE,
      caption = caption
    ) |>
      kable_styling(
        latex_options = c("hold_position", "scale_down"),
        font_size = font_size,
        full_width = FALSE
      )

    for (i in seq_along(widths)) {
      tbl <- tbl |> column_spec(i, width = widths[i])
    }

    print(tbl)

    if (landscape_pdf) {
      cat("\\end{landscape}\n")
    }
  } else {
    kable(
      data,
      format = "html",
      caption = caption,
      escape = FALSE
    ) |>
      kable_styling(
        full_width = FALSE,
        bootstrap_options = c("striped", "hover", "condensed")
      ) |>
      scroll_box(width = "100%", height = html_height)
  }
}
```

---

## Types of Consciousness

### By Philosophical Category

```{r taxonomy-types-consciousness, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
types_consciousness <- data.frame(
  Type = c(
    "Phenomenal consciousness",
    "Access consciousness",
    "Self-consciousness",
    "Reflective consciousness",
    "Pre-reflective consciousness",
    "Minimal consciousness",
    "Proto-consciousness",
    "Sentience",
    "Cosmic consciousness",
    "T-Consciousness",
    "Consciousness field",
    "Artificial consciousness"
  ),
  Definition = c(
    "Subjective experience; what it is like to see, feel, think, or exist from the first-person perspective",
    "Information available for reasoning, report, decision-making, and behavioural control",
    "Awareness of oneself as a subject or as being aware",
    "The capacity to think about one’s own thoughts, beliefs, identity, or experience",
    "Immediate lived awareness before explicit self-reflection",
    "Basic subjective awareness without complex self-reflection or language",
    "A possible precursor or minimal form of experience below full consciousness",
    "The capacity for felt experience, especially pleasure, pain, suffering, or well-being",
    "Universe-level or reality-level consciousness from which individual minds may derive",
    "Taheri’s proposed fundamental non-material consciousness reality that is not produced by matter, energy, frequency, or the brain",
    "A proposed non-material field-like organizing reality through which consciousness relates to or manifests matter, life, and mind",
    "Possible consciousness in artificial systems such as AI, robots, or artificial life"
  ),
  `Key theorists / frameworks` = c(
    "Nagel, Chalmers",
    "Block, Baars, Dehaene",
    "Rosenthal, higher-order theorists",
    "Higher-order theories, metacognition research",
    "Phenomenology, Merleau-Ponty",
    "Feinberg, Mallatt, minimal cognition debates",
    "Panpsychism, basal cognition, some evolutionary theories",
    "Bentham, Singer, animal ethics",
    "Cosmopsychism, some contemplative traditions",
    "Mohammad Ali Taheri",
    "Taheri’s T-Consciousness framework",
    "Functionalism, AI consciousness debates"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  types_consciousness,
  caption = "Types of consciousness by philosophical category.",
  widths = c("4.2cm", "11.5cm", "6.0cm"),
  font_size = 9,
  landscape_pdf = TRUE
)
```

This table shows that consciousness is not a single simple category. Some uses of the term refer to lived experience, while others refer to information access, self-reflection, sentience, cosmic awareness, artificial systems, or consciousness-first metaphysics.

---

## By Biological Level

```{r taxonomy-biological-levels, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
biological_levels <- data.frame(
  Level = c(
    "Molecular",
    "Cellular",
    "Unicellular organisms",
    "Multicellular non-neural life",
    "Simple nervous systems",
    "Invertebrates",
    "Vertebrates",
    "Primates and humans",
    "Artificial systems",
    "Artificial life"
  ),
  Examples = c(
    "Chemical reactions, proteins, microtubules",
    "Single cells, immune cells, developmental bioelectric networks",
    "Bacteria, protists, Paramecium, Stentor",
    "Plants, fungi, slime molds",
    "Cnidarians, jellyfish, hydra",
    "Insects, crustaceans, cephalopods",
    "Fish, amphibians, reptiles, birds, mammals",
    "Apes and humans",
    "AI, robots, artificial agents",
    "Digital organisms, synthetic cells, life-like simulations"
  ),
  `Evidence or argument` = c(
    "Speculative claims in Orch-OR and some quantum mind theories",
    "Sensing, memory, self-regulation, adaptive response",
    "Chemotaxis, habituation-like behaviour, decision-making, environmental responsiveness",
    "Signaling, learning-like behaviour, network optimization, resource allocation",
    "Nerve nets, sensory response, coordinated movement",
    "Learning, flexible behaviour, nociception, problem-solving, neural integration",
    "Central nervous systems, learning, pain behaviour, affect, perception",
    "Self-recognition, language, theory of mind, first-person reports",
    "Language, learning, self-modeling, goal-directed behaviour",
    "Evolution, adaptation, self-maintenance, reproduction-like processes"
  ),
  Status = c(
    "Highly speculative",
    "Cognition plausible; consciousness uncertain",
    "Basal cognition; consciousness debated",
    "Intelligence debated; consciousness uncertain",
    "Possible minimal sentience; uncertain",
    "Increasing evidence for sentience in some groups",
    "Strong evidence for many forms of consciousness",
    "Strongest evidence",
    "Unknown and debated",
    "Life-like status debated; consciousness uncertain"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  biological_levels,
  caption = "Biological and artificial levels relevant to consciousness.",
  widths = c("4.0cm", "5.3cm", "8.0cm", "5.3cm"),
  font_size = 8,
  landscape_pdf = TRUE
)
```

The biological taxonomy shows why the consciousness question becomes difficult at the lower levels of life. Evidence is strongest in humans and many vertebrates, increasingly plausible in some invertebrates, debated in plants and unicellular organisms, and highly speculative at the molecular level.

---

## By Ontological Level

```{r taxonomy-ontological-levels, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
ontological_levels <- data.frame(
  Level = c(
    "Brain-based consciousness",
    "Nervous-system consciousness",
    "Living-system consciousness",
    "Information-based consciousness",
    "Function-based consciousness",
    "Matter-based proto-consciousness",
    "Cosmos-based consciousness",
    "Consciousness-field ontology",
    "Mind-only ontology"
  ),
  Description = c(
    "Consciousness exists only in organisms with suitable brains",
    "Consciousness requires nervous systems but not necessarily human-like brains",
    "Consciousness or proto-consciousness may arise with living self-organization",
    "Consciousness depends on information integration or information architecture",
    "Consciousness depends on functional organization, regardless of substrate",
    "Matter has proto-conscious or experiential aspects",
    "The universe as a whole is the primary conscious subject",
    "Consciousness is a fundamental non-material reality expressed through consciousness fields",
    "Consciousness or mind is the only ultimate reality"
  ),
  `Example theories` = c(
    "Biological naturalism, GWT, RPT",
    "RPT, neurobiological theories, animal consciousness frameworks",
    "FEP, autopoietic enactivism, co-emergence",
    "IIT, computational theories",
    "Functionalism, AST, GWT-inspired AI theories",
    "Panpsychism, Russellian monism",
    "Cosmopsychism",
    "Taheri’s T-Consciousness framework",
    "Idealism, analytic idealism"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  ontological_levels,
  caption = "Ontological levels at which consciousness may be located.",
  widths = c("5.2cm", "10.5cm", "6.5cm"),
  font_size = 9,
  landscape_pdf = TRUE
)
```

This table helps distinguish theories that locate consciousness in brains, nervous systems, life, information, function, matter, the cosmos, consciousness fields, or mind itself.

---

## Levels and Dimensions

### Graded Dimensions of Consciousness

```{r taxonomy-graded-dimensions, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
graded_dimensions <- data.frame(
  Dimension = c(
    "Wakefulness",
    "Awareness",
    "Sentience",
    "Self-awareness",
    "Bodily awareness",
    "Temporal depth",
    "Intentional complexity",
    "Integration",
    "Reportability",
    "Agency",
    "Meaning"
  ),
  `Lower end` = c(
    "Deep sleep, coma, anaesthesia",
    "No apparent awareness",
    "No evidence of feeling",
    "No self-model",
    "Minimal bodily regulation",
    "Immediate reaction",
    "Simple world-directedness",
    "Fragmented processing",
    "No report possible",
    "Passive response",
    "Bare signal response"
  ),
  `Higher end` = c(
    "Alert waking state",
    "Rich phenomenal experience",
    "Pain, pleasure, suffering, well-being",
    "Reflective awareness of oneself as a subject",
    "Rich embodied self-experience",
    "Memory, anticipation, planning",
    "Nested beliefs, theory of mind, symbolic thought",
    "Unified field of experience",
    "Verbal or symbolic report",
    "Flexible, self-directed action",
    "Organism-relative significance and interpretation"
  ),
  Relevance = c(
    "Measures arousal, not necessarily awareness",
    "Core dimension of consciousness",
    "Central for moral status",
    "Important in humans and some animals",
    "Links consciousness to life and organismic regulation",
    "Important for complex cognition",
    "Important for higher cognition",
    "Important in IIT, GWT, and neurobiological theories",
    "Useful but not identical to consciousness",
    "Bridges life, cognition, and consciousness",
    "Central to biosemiotics and co-emergence"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  graded_dimensions,
  caption = "Graded dimensions of consciousness.",
  widths = c("4.0cm", "5.2cm", "6.7cm", "6.2cm"),
  font_size = 8,
  landscape_pdf = TRUE
)
```

Consciousness is not only present or absent. It varies across dimensions such as wakefulness, awareness, selfhood, agency, integration, reportability, and meaning. This is why a graded model is often more useful than a simple yes-or-no model.

---

## States of Consciousness

```{r taxonomy-states-consciousness, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
states_consciousness <- data.frame(
  State = c(
    "Deep sleep",
    "Dreaming",
    "Lucid dreaming",
    "Waking state",
    "Flow state",
    "Meditation",
    "Anaesthesia",
    "Coma",
    "Vegetative state / unresponsive wakefulness",
    "Minimally conscious state",
    "Locked-in syndrome",
    "Psychedelic states",
    "Dissociative states",
    "Near-death experiences"
  ),
  Wakefulness = c(
    "Low",
    "Low to moderate",
    "Low to moderate",
    "High",
    "High",
    "Variable",
    "Very low or absent",
    "Very low or absent",
    "Wakefulness cycles may remain",
    "Variable",
    "High",
    "High or altered",
    "Variable",
    "Variable / medically complex"
  ),
  Awareness = c(
    "Low or absent",
    "Moderate to high",
    "High",
    "High",
    "High",
    "Often high",
    "Absent or severely reduced",
    "None or minimal",
    "No clear behavioural awareness",
    "Partial or inconsistent",
    "High",
    "High and altered",
    "Variable",
    "Reported as vivid by some individuals"
  ),
  `Self-awareness` = c(
    "None",
    "Variable",
    "High",
    "Variable to high",
    "Often reduced",
    "Variable",
    "None",
    "None",
    "None apparent",
    "Limited",
    "High",
    "Variable",
    "Altered",
    "Often high in reports"
  ),
  Notes = c(
    "Usually low consciousness, though some dreaming may occur in sleep states",
    "Rich experience can occur without ordinary waking control",
    "Dreamer recognizes the dream state",
    "Ordinary conscious life",
    "Strong task absorption with reduced reflective self-consciousness",
    "May involve altered attention, selfhood, or awareness",
    "Used to study loss and recovery of consciousness",
    "Severe impairment of wakefulness and awareness",
    "Raises difficult diagnostic and ethical issues",
    "Some evidence of awareness remains",
    "Consciousness preserved but motor output severely limited",
    "May involve changes in self, perception, and meaning",
    "Self-experience and embodiment may fragment",
    "Interpretation remains debated"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  states_consciousness,
  caption = "States of consciousness.",
  widths = c("4.5cm", "3.7cm", "4.4cm", "3.6cm", "7.5cm"),
  font_size = 8,
  landscape_pdf = TRUE,
  html_height = "600px"
)
```

States of consciousness show that wakefulness and awareness are not the same. A dreaming person may have vivid experience with reduced wakefulness, while a locked-in patient may have high awareness but little or no motor expression.

---

## Theory Classification

### By Ontological Commitment

```{r taxonomy-ontological-commitment, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
ontology_commitment <- data.frame(
  Category = c(
    "Physicalist",
    "Biological naturalist",
    "Functionalist",
    "Informational",
    "Enactive / autopoietic",
    "Panpsychist",
    "Cosmopsychist",
    "Idealist",
    "T-Consciousness framework",
    "Dualist",
    "Neutral monist",
    "Dual-aspect monist",
    "Process philosophical",
    "Quantum speculative"
  ),
  Theories = c(
    "GWT, RPT, AST, emergentism, many neuroscientific theories",
    "Searle, neurobiological theories",
    "Computational functionalism, some AI consciousness theories",
    "IIT, some information-based theories",
    "Varela, Thompson, Di Paolo",
    "Strawson, Goff, some readings of Russellian monism",
    "Shani, Nagasawa, Goff",
    "Berkeley, Kastrup, analytic idealism",
    "Taheri’s consciousness-field ontology",
    "Substance dualism, property dualism",
    "Russell, James, some dual-aspect views",
    "Spinoza, Pauli-Jung interpretations",
    "Whitehead",
    "Orch-OR, Stapp, Fisher, Bohm-inspired theories"
  ),
  `Core claim` = c(
    "Consciousness arises from physical processes",
    "Consciousness is a biological phenomenon caused by brain processes",
    "Consciousness depends on functional organization",
    "Consciousness depends on integrated information or causal structure",
    "Mind arises through living self-organization and embodied sense-making",
    "Consciousness or proto-consciousness is fundamental and widespread",
    "The universe as a whole is conscious; individuals derive from it",
    "Consciousness is the only ultimate reality",
    "T-Consciousness is fundamental, non-material, and not produced by the brain",
    "Mind and matter are distinct substances or properties",
    "Mind and matter arise from a deeper neutral reality",
    "One reality has mental and physical aspects",
    "Reality is composed of processes or occasions of experience",
    "Consciousness involves quantum or deeper physical processes"
  ),
  `Relation to life-consciousness` = c(
    "Life first",
    "Life first",
    "Potentially substrate-independent",
    "Co-emergence or panpsychist implications",
    "Co-emergence",
    "Consciousness first or co-emergence",
    "Consciousness first",
    "Consciousness first",
    "Consciousness first",
    "Separates life and consciousness",
    "Co-emergence",
    "Co-emergence",
    "Co-emergence or consciousness first",
    "Co-emergence or consciousness first"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  ontology_commitment,
  caption = "Theories classified by ontological commitment.",
  widths = c("4.0cm", "6.5cm", "8.0cm", "5.0cm"),
  font_size = 7,
  landscape_pdf = TRUE,
  html_height = "600px"
)
```

This classification shows that theories of consciousness often differ less in evidence than in starting assumptions. A physicalist, an idealist, and a T-Consciousness theorist may all discuss consciousness, but they locate it at very different levels of reality.

---

## By Mechanism

```{r taxonomy-mechanisms, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
mechanism_table <- data.frame(
  `Mechanism type` = c(
    "Global access",
    "Recurrent processing",
    "Higher-order monitoring",
    "Attention modeling",
    "Information integration",
    "Predictive regulation",
    "Biological embodiment",
    "Self-production",
    "Semiotic meaning",
    "Quantum physical",
    "Consciousness-field organization",
    "Philosophical grounding"
  ),
  Theories = c(
    "GWT, Global Neuronal Workspace",
    "RPT, Edelman’s re-entry",
    "Higher-order thought and perception theories",
    "AST",
    "IIT",
    "Predictive processing, FEP",
    "Biological naturalism, Damasio, enactivism",
    "Autopoiesis",
    "Biosemiotics",
    "Orch-OR, Stapp, Fisher",
    "Taheri’s T-Consciousness framework",
    "Panpsychism, idealism, cosmopsychism"
  ),
  `Key mechanism` = c(
    "Information broadcast across a cognitive system",
    "Neural feedback loops",
    "A mental state becomes conscious when represented by another state",
    "The system models its own attention",
    "Integrated causal information, Phi",
    "Prediction-error minimization and active inference",
    "Living bodies, affect, regulation, and neural processes",
    "The system continuously produces and maintains itself",
    "Sign interpretation and biological meaning",
    "Quantum coherence, collapse, or measurement",
    "Non-material consciousness fields organize or manifest matter, life, and mind",
    "Consciousness as fundamental reality"
  ),
  `Main question` = c(
    "Does access explain experience?",
    "Is recurrence sufficient for consciousness?",
    "Does consciousness require self-monitoring?",
    "Does a model of awareness create awareness?",
    "Can consciousness be mathematically measured?",
    "Does prediction become experience?",
    "Does consciousness require life?",
    "Is life already proto-cognitive?",
    "Does life require meaning?",
    "Does quantum physics explain experience?",
    "Can this be connected to testable mechanisms?",
    "Does this explain or relocate the mystery?"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  mechanism_table,
  caption = "Theories classified by proposed mechanism.",
  widths = c("5.0cm", "5.5cm", "8.0cm", "5.0cm"),
  font_size = 8,
  landscape_pdf = TRUE
)
```

Mechanism-based classification is useful because some theories are scientific models while others are metaphysical frameworks. Global Workspace Theory, Recurrent Processing Theory, and Attention Schema Theory propose mechanisms inside cognitive systems. Idealism, panpsychism, and Taheri’s T-Consciousness framework begin from broader claims about the nature of reality.

---

## By Relation to Life

```{r taxonomy-relation-to-life, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
relation_life <- data.frame(
  View = c(
    "Strict life-first",
    "Neural life-first",
    "Biological naturalism",
    "Co-emergence",
    "Autopoietic-enactive view",
    "Panpsychism",
    "Cosmopsychism",
    "Idealism",
    "Taheri’s T-Consciousness framework",
    "Artificial consciousness functionalism",
    "Artificial life view"
  ),
  `Life first?` = c(
    "Yes",
    "Yes",
    "Yes",
    "Partly",
    "Life begins cognition",
    "No",
    "No",
    "No",
    "No",
    "Not necessarily",
    "Artificial life may come first"
  ),
  `Consciousness first?` = c(
    "No",
    "No",
    "No",
    "Partly",
    "Not necessarily",
    "Yes or always present",
    "Yes",
    "Yes",
    "Yes",
    "No or depends",
    "Possibly"
  ),
  `Core interpretation` = c(
    "Consciousness is a late product of biological evolution",
    "Consciousness appears only with nervous systems or brains",
    "Consciousness is caused by biological brain processes",
    "Life and consciousness are deeply linked through self-organization and meaning",
    "Mind is rooted in living self-production",
    "Consciousness or proto-consciousness is fundamental in matter",
    "Cosmic consciousness precedes individual life",
    "Life appears within consciousness",
    "Life manifests through or within T-Consciousness and consciousness fields",
    "Consciousness may arise in non-living functional systems",
    "Consciousness may arise in life-like artificial systems"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  relation_life,
  caption = "Theories classified by relation between life and consciousness.",
  widths = c("5.0cm", "3.5cm", "4.0cm", "10.0cm"),
  font_size = 8,
  landscape_pdf = TRUE
)
```

This table directly connects the taxonomy to the central question of the book. Some theories place life before consciousness. Others place consciousness before life. Co-emergence views suggest that the relationship may not be strictly sequential.

---

## By Evidence Type

```{r taxonomy-evidence-type, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
evidence_type <- data.frame(
  `Evidence type` = c(
    "First-person report",
    "Behaviour",
    "Neural activity",
    "Brain stimulation / lesion evidence",
    "Anaesthesia response",
    "Evolutionary continuity",
    "Basal cognition evidence",
    "Formal measures",
    "Artificial system behaviour",
    "Contemplative / phenomenological data",
    "Ethical precaution"
  ),
  `Most relevant to` = c(
    "Human consciousness",
    "Animals, infants, AI, simple organisms",
    "Humans and animals with nervous systems",
    "Human and animal consciousness",
    "Humans, animals, some biological systems",
    "Animal consciousness",
    "Simple organisms and plants",
    "IIT, computational theories",
    "AI consciousness debates",
    "Human experience",
    "Animal and AI moral status"
  ),
  Strength = c(
    "Direct access to reported experience",
    "Widely applicable",
    "Strong correlation with conscious states",
    "Helps identify causal mechanisms",
    "Useful for studying loss of consciousness",
    "Supports graded approaches",
    "Shows intelligence-like life processes",
    "Provides precision",
    "Tests substrate independence",
    "Rich first-person detail",
    "Useful under uncertainty"
  ),
  Limitation = c(
    "Limited to beings that can report",
    "Behaviour may be unconscious",
    "Correlation is not explanation",
    "Mostly limited to nervous systems",
    "Does not define consciousness by itself",
    "Does not identify exact thresholds",
    "Does not prove experience",
    "May not capture phenomenology",
    "Intelligence may not equal sentience",
    "Hard to standardize",
    "May overextend moral concern"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  evidence_type,
  caption = "Evidence types used in consciousness research.",
  widths = c("5.0cm", "6.0cm", "6.0cm", "6.0cm"),
  font_size = 8,
  landscape_pdf = TRUE
)
```

Different kinds of evidence answer different questions. First-person reports are powerful but limited to beings who can report. Behaviour is widely available but may not prove experience. Neural evidence is strong in animals with nervous systems but does not apply easily to plants, microbes, or artificial systems. Formal measures offer precision but may miss lived experience.

---

## Summary of Main Taxonomic Distinctions

```{r taxonomy-main-distinctions, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
main_distinctions <- data.frame(
  Distinction = c(
    "Phenomenal vs access consciousness",
    "Cognition vs consciousness",
    "Sentience vs intelligence",
    "Self-awareness vs consciousness",
    "Life vs mind",
    "Simulation vs realization",
    "Weak vs strong emergence",
    "Matter-based vs consciousness-based ontology",
    "Bottom-up vs top-down explanation",
    "Scientific vs metaphysical theory",
    "Consciousness field vs physical field"
  ),
  Meaning = c(
    "Experience itself vs information available for use",
    "Information processing vs subjective awareness",
    "Capacity to feel vs capacity to solve problems",
    "Reflective self-model vs basic experience",
    "Biological organization vs experience or cognition",
    "Imitating a process vs actually instantiating it",
    "Reducible higher-level pattern vs irreducible novelty",
    "Matter as primary vs consciousness as primary",
    "Parts produce wholes vs wholes or fields constrain parts",
    "Empirically testable model vs fundamental worldview",
    "Non-material organizing reality vs measurable physical field"
  ),
  `Why it matters` = c(
    "A system may use information without experience",
    "Simple organisms may be cognitive but not conscious",
    "Moral status depends more on sentience than intelligence",
    "Many beings may be conscious without reflective selfhood",
    "Central to the book’s main question",
    "Crucial for AI and artificial life",
    "Central to emergentist theories",
    "Determines whether life or consciousness comes first",
    "Important in origin-of-life and consciousness-first models",
    "Important for evaluating testability",
    "Important for placing T-Consciousness correctly"
  ),
  check.names = FALSE
)

render_taxonomy_table(
  main_distinctions,
  caption = "Summary of main taxonomic distinctions.",
  widths = c("6.5cm", "7.5cm", "7.5cm"),
  font_size = 9,
  landscape_pdf = TRUE
)
```

---

## Taxonomic Caution

This taxonomy should be read as a map, not a final theory. Different traditions divide consciousness in different ways. Some theories treat consciousness as neural access. Others treat it as subjective experience. Others treat it as integrated information, biological meaning, cosmic awareness, or a non-material consciousness field.

The same term may therefore have different meanings depending on context. For example, “consciousness” in Global Workspace Theory does not mean exactly the same thing as “consciousness” in idealism, panpsychism, or Taheri’s T-Consciousness framework. Similarly, “information” in Shannon theory, biology, IIT, and biosemiotics does not always mean the same thing.

The purpose of this taxonomy is to make these differences visible. It helps readers see that disagreement about consciousness is often not only disagreement about evidence. It is also disagreement about classification, definition, and metaphysical starting point.

